Live Functional Programming with Typed Holes

### 2018-02-09 abstract.tex
wrapping and separating lines; no changes to text
edits to abstract

\begin{abstract}
\emph{Live programming environments} aim to provide programmers (and, in some cases, audiences) with continuous feedback about a program's dynamic behavior throughout the editing process. 
The problem that motivates this paper is that programming languages typically assign dynamic meaning only to programs that are syntactically well-formed and free of static type errors (collectively, \emph{complete programs}), whereas program editors frequently encounter incomplete edit states. 
Recent work by \citet{popl-paper} has made progress on the problem of working directly with incomplete programs by developing a type system for lambda terms with \emph{holes} in both expression and type position, with empty holes standing for missing expressions or types, and non-empty holes operating as ``membranes'' around type inconsistencies. 
This paper develops a corresponding dynamic semantics for this class of incomplete functional programs. 
Rather than aborting with an exception when evaluation encounters a hole (as in  several existing systems), evaluation proceeds ``around'' expression holes, performing as much computation as is possible and tracking the dynamic environment around each hole as it flows through the program. 
These accumulated hole environments can be reported directly to the programmer by the editor as they work to fill holes. 
They also enable an ``edit-and-resume'' feature that avoids the need to restart evaluation after edits that amount to hole filling. 
Formally, our semantics combines features from gradual type theory (to handle incomplete types) and contextual modal type theory (CMTT, which provides a logical foundation for hole environment tracking) with novel structure necessary to prove key metatheoretic properties, including 1) a type safety property that accounts for expression holes (i.e. free metavariables, from the perspective of CMTT), and 2) commutativity and confluence properties that guarantee that the ``edit-and-resume'' feature is sound. 
The primary focus of this paper is on this formal development, which we have mechanized using the Agda proof assistant. 
We additionally describe a simple reference implementation, called Hazelnut Live, that exposes a language of structured edit actions that insert holes as necessary to guarantee that every edit state has some (possibly incomplete) type, based on the edit action calculus in the prior work by \citet{popl-paper}. 
Taken together, the result is a proof-of-concept live functional programming environment where dynamic feedback is truly continuous, i.e. it is available for every possible edit state. % i.e. \emph{every edit state is guaranteed to have non-trivial static and dynamic meaning}.

>>>>

\begin{abstract}

\emph{Live programming environments} aim to provide programmers (and, in some
cases, audiences) with continuous feedback about a program's dynamic behavior
throughout the editing process.
%
Unfortunately, because programming languages typically assign dynamic meaning
only to programs that are \emph{complete}---syntactically well-formed and free
of static type errors---live programming features are often severely restricted,
if available at all, when programs are incomplete.
%
Incomplete edit states are frequently encountered when writing programs, so they
are, arguably, when live programming services would be most valuable.
%
Defining a semantics for incomplete functional programs would help bridge the
gap between well-formed states and allowing live programming environments to
provide continuous, uninterrupted, and meaningful feedback.

Recent work by \citet{popl-paper} makes progress toward this goal by developing
a type system for lambda terms with \emph{holes} in both expression and type
position, with empty holes standing for missing expressions or types, and
non-empty holes operating as ``membranes'' around type inconsistencies.
%
The contribution of this paper is a corresponding \emph{dynamic semantics for
functional programs with holes}.
%
Rather than aborting with an exception when evaluation encounters a hole (as in
several existing systems), evaluation proceeds ``around'' expression holes,
performing as much computation as is possible and tracking the dynamic
environment around each hole as it flows through the program.  These accumulated
hole environments can be reported directly to the programmer by the editor as
they work to fill holes.  They also enable an ``edit-and-resume'' feature that
avoids the need to restart evaluation after edits that amount to hole filling.

Formally, our semantics combines features from gradual type theory (to handle
incomplete types) and contextual modal type theory (CMTT, which provides a
logical foundation for hole environment tracking) with novel structure necessary
to prove key metatheoretic properties, including (1) a type safety property that
accounts for expression holes (i.e. free metavariables, from the perspective of
CMTT), and (2) commutativity and confluence properties that guarantee that the
``edit-and-resume'' feature is sound.
%
We have mechanized our formal development using the Agda proof assistant.
%
We additionally describe a simple reference implementation, called Hazelnut
Live, that exposes a language of structured edit actions that insert holes as
necessary to guarantee that every edit state has some (possibly incomplete)
type, based on the edit action calculus in the prior work by \citet{popl-paper}.
%
Taken together, the result is a proof-of-concept live functional programming
environment where dynamic feedback is truly continuous, i.e. it is available for
every possible edit state. % i.e. \emph{every edit state is guaranteed to have
non-trivial static and dynamic meaning}.
